Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package \vskip15pt Supplementary Materials

### 2018-07-13 docs/MEE_submission/manuscript_MEE.tex
final MS updates

\section{\pkg{metaDigitise} and Reproducibility}
The \pkg{metaDigitise} package has one main function, \fct{metaDigitise}, which interactively takes the user through the process of extracting data from figures. \fct{metaDigitise} works on a directory containing images of figures copied from primary literature, in .png, .jpg, .tiff, .pdf format, specified to \fct{metaDigitise} through the \code{dir} argument. \fct{metaDigitise} recognizes all the images in the given directory and automatically imports them one-by-one, allowing the user to extract the relevant information about a figure as they go. Figures can be organised in different ways for a project, but we would recommend having all figures for one project in a single directory with an informative and unambiguous naming scheme (e.g. paper\_figure\_trait.png). This expedites digitisation by preventing users from having to constantly change directories and / or open new images.

@@

As figures may have been copied from older, scanned publications, they may not be perfectly orientated. This makes calibration of the points in the figure problematic. \fct{metaDigitise} allows users to rotate the image (Figure S2A,B). Furthermore, mean/error plots, box plots and histograms, may be presented with horizontal bars. \fct{metaDigitise} assumes that bars are vertical, but allows the user to flip the image to make the bars are vertical (Figure S2C,D).

%%% \pkg{metaDigitise} also allows back calculation of data presented on log axes.

@@

\begin{figure}[!h]
\centering 
\rotatebox{90}{%

 \includegraphics[width=1.3\textwidth]{Adobe_fig1_V3.png} 
 }
 \caption{Functionality of \pkg{metaDigitise}. Using the iris dataset in R, digitisation of different plot types, A) mean/error plot, B) box plot, C) histogram and D) scatter plot, is shown in \pkg{metaDigitise} (left) compared with other common softwares (right). A) and B) are plotted with the whole dataset, C) is just the data for the species \textit{setosa} and D) a subset from all three species. Notable functions of metaDigitise are listed in the center. Other software also perform points 3 and 4 (see Table \ref{tab:comparison}), although these functions are more developed in \pkg{metaDigitise}. As shown on the left hand side of the figure, \pkg{metaDigitise} clearly displays the stages of the digitisation to aid the transparency of the process, and returns concatenated summary data for all images.
%  All figures are clearly labelled at the top to remind users of the filename and plot type. This reduces errors throughout the digitisation process. Names of the variables and calibration (in blue) are plotted alongside the digitised points. In A) and B), user entered group names and sample sizes are displayed beside the relevant points. In C) the names and sample sizes for each group are shown below the figure.
}
\label{fig:all_extract}
\end{figure}

>>>>

\section{\pkg{metaDigitise} and Reproducibility}
The \pkg{metaDigitise} package has one main function, \fct{metaDigitise}, which interactively takes the user through the process of extracting data from figures (see Supplementary Material S1 for a full tutorial). \fct{metaDigitise} works on a directory containing images of figures copied from primary literature, in .png, .jpg, .tiff, .pdf format, specified to \fct{metaDigitise} through the \code{dir} argument. \fct{metaDigitise} recognizes all the images in the given directory and automatically imports them one-by-one, allowing the user to extract the relevant information about a figure as they go. Figures can be organised in different ways for a project, but we would recommend having all figures for one project in a single directory with an informative and unambiguous naming scheme (e.g. paper\_figure\_trait.png). This expedites digitisation by preventing users from having to constantly change directories and / or open new images.

@@

As figures may have been copied from older, scanned publications, they may not be perfectly orientated. This makes calibration of the points in the figure problematic. \fct{metaDigitise} allows users to rotate the image (Figure S2A,B). Furthermore, mean/error plots, box plots and histograms, may be presented with horizontal bars. \fct{metaDigitise} assumes that bars are vertical, but allows the user to flip the image to make the bars are vertical (Figure S2C,D). \pkg{metaDigitise} also allows back calculation of data presented on log axes.

@@

\begin{figure}[!h]
\centering 
% \rotatebox{90}{%

%  \includegraphics[width=1.3\textwidth]{Adobe_fig1_V3.png} 
%  }
 \caption{\doublespacing
Functionality of \pkg{metaDigitise}. Using the iris dataset in R, digitisation of different plot types, A) mean/error plot, B) box plot, C) histogram and D) scatter plot, is shown in \pkg{metaDigitise} (left) compared with other common softwares (right). A) and B) are plotted with the whole dataset, C) is just the data for the species \textit{setosa} and D) a subset from all three species. Notable functions of metaDigitise are listed in the center. Other software also perform points 3 and 4 (see Table \ref{tab:comparison}), although these functions are more developed in \pkg{metaDigitise}. As shown on the left hand side of the figure, \pkg{metaDigitise} clearly displays the stages of the digitisation to aid the transparency of the process, and returns concatenated summary data for all images.
%  All figures are clearly labelled at the top to remind users of the filename and plot type. This reduces errors throughout the digitisation process. Names of the variables and calibration (in blue) are plotted alongside the digitised points. In A) and B), user entered group names and sample sizes are displayed beside the relevant points. In C) the names and sample sizes for each group are shown below the figure.
}
\label{fig:all_extract}
\end{figure}

### 2018-07-13 docs/MEE_submission/supplements.tex
final MS updates

\subsection{Inter-observer variability in digitisations}
In order to evaluate the consistency of digitisation using \pkg{metaDigitise} between users, we simulated a dataset of two variables with two groups ($n=10$ within groups). Each variable was plotted twice for each plot type (figures were modified slightly to give users a sense that they were digitising new data) generating a total of 14 figures. Fourteen independent digitisers (including the authors) were provided with a directory with all 14 figures in a randomised order. Digitisers ran \pkg{metaDigistise} on their own computers, across different operating systems (including Mac, Windows and Linux). Digitisers varied in their level of experience, from people with experience of meta-analyses to those without any science background. We asked users to digitise all 14 figures and collected the mean, standard deviation and correlation coefficient ($r$, for scatter plots) generated by \fct{metaDigitise} for every plot digitised ($n=28$ per digitiser per metric, $n=4$ for $r$).

@@

where $\theta$ is the estimate and $\hat{\theta}$ is the true value. The deviation from the true value of $r$ was not further standardised, as it is already on a standardised scale. We also took the absolute values of these standardised differences as a measure of precision. The resulting data was used to assess between- and within- user variability (i.e., the intra-class correlation coefficient - ICC). This was done using linear mixed effect models with user identity as a random effect using \pkg{lme4} \citep{bates2015} in \proglang{R}. Standardised mean, standard deviation and correlation coefficients were used as response variables in separate models. Sampling variance for ICC estimates was generated based on 1000 parametric bootstraps of the model and the significance was tested using likelihood ratio tests, using \pkg{rptR} \citep{stoffel2017}.

>>>>

\subsection{Inter-observer variability in digitisations}
In order to evaluate the consistency of digitisation using \pkg{metaDigitise} between users, we simulated a dataset of two variables with two groups ($n=10$ within groups). Each variable was plotted twice for each plot type (figures were modified slightly to give users a sense that they were digitising new data) generating a total of 14 figures (example figures are available at http://doi.org/10.5281/zenodo.1311681). Fourteen independent digitisers (including the authors) were provided with a directory with all 14 figures in a randomised order. Digitisers ran \pkg{metaDigistise} on their own computers, across different operating systems (including Mac, Windows and Linux). Digitisers varied in their level of experience, from people with experience of meta-analyses to those without any science background. We asked users to digitise all 14 figures and collected the mean, standard deviation and correlation coefficient ($r$, for scatter plots) generated by \fct{metaDigitise} for every plot digitised ($n=28$ per digitiser per metric, $n=4$ for $r$).

@@

where $\theta$ is the estimate and $\hat{\theta}$ is the true value. The deviation from the true value of $r$ was not further standardised, as it is already on a standardised scale. We also took the absolute values of these standardised differences as a measure of precision. The resulting data was used to assess between- and within- user variability (i.e., the intra-class correlation coefficient - ICC). This was done using linear mixed effect models with user identity as a random effect using \pkg{lme4} \citep{bates2015} in \proglang{R}. Standardised mean, standard deviation and correlation coefficients were used as response variables in separate models. Sampling variance for ICC estimates was generated based on 1000 parametric bootstraps of the model and the significance was tested using likelihood ratio tests, using \pkg{rptR} \citep{stoffel2017}. Data and code are available at http://doi.org/10.5281/zenodo.1311681.
